---
title: "lexis Demo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lexis Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`lexis` provides word-level psycholinguistic norms and helper functions for
lookup, filtering, definitions, embeddings, and matching.

## Load package

```{r}
library(lexis)
```

## Quick lookup

Use `lex_lookup()` to retrieve values for one or more words.

```{r}
lex_lookup(c("dog", "justice"), dims = c("aoa", "concreteness", "valence"))
```

Use `wide = TRUE` for one row per word.

```{r}
lex_lookup(c("dog", "justice"), wide = TRUE)
```

## Search and filter

`lex_search()` helps find words by pattern.

```{r}
lex_search("^bio", cols = "word")
```

`lex_filter()` filters on numeric dimensions.

```{r}
lex_filter(concreteness > 4, aoa < 8)
```

## Dictionary definitions

Use `lex_define()` to pull Wordset definitions.

```{r}
lex_define(c("bank", "charge"), n = 3)
```

## Embeddings

Get vectors with `lex_embed()`.

```{r}
emb <- lex_embed(c("dog", "cat", "wolf"))
dim(emb)
```

Nearest neighbors with `lex_neighbors()`.

```{r}
lex_neighbors("dog", n = 10)
```

## Stimulus matching

`lex_match()` finds words matched on selected dimensions.

```{r}
lex_match(
  targets = c("anger", "joy"),
  dims = c("aoa", "concreteness", "valence"),
  n = 3
)
```
