---
title: "LLM Norming Pipeline (sampled Monte-Carlo expected value, direct OpenAI API)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

## Purpose

A logprob-free version of the distributional ("expected value") extraction. Instead
of reading `top_logprobs` to compute Σ rating·p(rating) analytically, we draw `k`
integer ratings per word at temperature 1.0 and take their **mean** — a Monte-Carlo
estimate of the same quantity. This works on any model, including those that never
expose logprobs (Claude, gpt-5 series), so it tests whether the expected-value
benefit established on the gpt-4 logprob models is **recoverable without logprobs**.

This run targets `gpt-4.1` and `gpt-4o-mini` — the models where the analytic logprob
`expected` value already exists — so the sampled mean can be validated directly
against it.

**Why a direct API call (not ellmer):** the `k` draws are requested in a single call
via the OpenAI `n` parameter, which returns `n` completions in one response with the
prompt billed **once**. ellmer is built around single-turn chats and surfaces only
the first choice, so it cannot return the other draws — hence raw httr2 here.

Output schema matches the other runs (`expected`, `integer_rating`, `coverage`) so
the run flows through the multi-model analysis loader. `extraction` is `"sampled"`;
per-task summaries land in `tasks/` (19 files), and per-draw ratings go to a
top-level `draws.csv` (analogous to the logprob `distributions.csv`).

## Run Controls

```{r controls}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(purrr)
  library(jsonlite)
  library(httr2)
})

# ── Pick model for this run ───────────────────────────────────────────────────
model_slug <- "gpt-4o-mini"   # or "gpt-4.1"

n_draws     <- 20L            # k: draws per word (matches top_logprobs=20 budget)
sample_temp <- 1.0            # native output distribution

MODELS <- list(
  "gpt-4.1"     = list(api_model = "gpt-4.1",     api_key_env = "OPENAI_API_KEY"),
  "gpt-4o-mini" = list(api_model = "gpt-4o-mini", api_key_env = "OPENAI_API_KEY")
)

model_cfg <- MODELS[[model_slug]]
if (is.null(model_cfg)) {
  stop("Unknown model_slug: ", model_slug,
       ". Choose one of: ", paste(names(MODELS), collapse = ", "), call. = FALSE)
}

api_url     <- "https://api.openai.com/v1/chat/completions"
max_tokens  <- 16L
max_active  <- 8L             # concurrent in-flight requests

# Cost lever (safe). The equivalence test does not need all ~1,638 holdout
# words/dimension — a few hundred gives a tight estimate. Per-dimension cap;
# NULL = all. The existing logprob run covers all holdout words, so any subset
# still joins cleanly for validation.
words_per_dim_cap <- NULL   # NULL = full holdout (no cap)
sample_seed       <- 1L

norm_samples_path     <- "../output/core_vocab_long.csv"
instructions_rda_path <- "../../data/norming_instructions.rda"
output_root           <- "../output/runs"

exclude_constructs <- c("lexical_decision_reaction_time",
                        "speeded_naming_reaction_time")
include_constructs <- character(0)
```

## Helpers

```{r helpers}
ensure_dir <- function(path) {
  dir.create(path, recursive = TRUE, showWarnings = FALSE)
  invisible(path)
}

read_norming_instructions <- function(path) {
  if (!file.exists(path)) stop("File not found: ", path, call. = FALSE)
  env <- new.env(parent = emptyenv())
  load(path, envir = env)
  get("norming_instructions", envir = env, inherits = FALSE)
}

build_instruction_map <- function(instructions_df) {
  instructions_df |>
    transmute(dataset = .data$dataset, construct_name = .data$construct_name,
              scale_min = .data$scale_min, scale_max = .data$scale_max,
              instructions_llm = .data$instructions_llm)
}

build_tasks <- function(instruction_map, norm_samples,
                        include_constructs, exclude_constructs) {
  tasks <- instruction_map |>
    distinct(.data$dataset, .data$construct_name, .data$scale_min, .data$scale_max) |>
    filter(!(.data$construct_name %in% exclude_constructs)) |>
    mutate(dimension = .data$construct_name) |>
    filter(.data$dimension %in% norm_samples$dimension)

  if (length(include_constructs)) {
    tasks <- filter(tasks, .data$construct_name %in% include_constructs)
  }

  tasks |>
    transmute(task_id = paste0(.data$construct_name, "_sampled"),
              dataset = .data$dataset, dimension = .data$dimension,
              scale_min = .data$scale_min, scale_max = .data$scale_max)
}

build_prompts <- function(task_rows, norm_samples, instruction_map, max_words,
                          words_cap = NULL, seed = 1L) {
  bind_rows(lapply(seq_len(nrow(task_rows)), function(i) {
    task <- as.list(task_rows[i, , drop = FALSE])
    instr <- instruction_map |>
      filter(.data$dataset == task$dataset, .data$construct_name == task$dimension) |>
      pull(.data$instructions_llm) |>
      dplyr::first()
    words <- norm_samples |>
      filter(.data$dimension == task$dimension, .data$split == "holdout") |>
      slice_head(n = max_words) |>
      select("word", "score", "split")
    if (!is.null(words_cap) && nrow(words) > words_cap) {
      set.seed(seed + i)              # reproducible, distinct per dimension
      words <- slice_sample(words, n = words_cap)
    }
    crossing(task_rows[i, c("task_id", "dataset", "dimension",
                            "scale_min", "scale_max")], words) |>
      mutate(instruction_text = instr)
  }))
}

parse_integer_text <- function(raw, scale_min, scale_max) {
  if (is.null(raw) || is.na(raw) || !nzchar(raw)) return(NA_integer_)
  raw <- trimws(raw)
  val <- suppressWarnings(as.integer(raw))
  if (!is.na(val) && val >= scale_min && val <= scale_max) return(val)
  hits <- regmatches(raw, gregexpr("[0-9]+", raw))[[1]]
  if (!length(hits)) return(NA_integer_)
  vals <- as.integer(hits)
  in_range <- vals[vals >= scale_min & vals <= scale_max]
  if (length(in_range)) return(in_range[[1]])
  NA_integer_
}

# mode of a numeric vector (lowest value on ties); the sampled integer point
# estimate, analogous to the logprob argmax.
sample_mode <- function(x) {
  x <- x[!is.na(x)]
  if (!length(x)) return(NA_integer_)
  tb <- sort(table(x), decreasing = TRUE)
  as.integer(names(tb)[tb == max(tb)][1])
}

# One chat/completions request returning n_draws completions for a single word.
build_req <- function(prompt_text, api_url, api_model, api_key,
                      n, temperature, max_tokens) {
  request(api_url) |>
    req_auth_bearer_token(api_key) |>
    req_body_json(list(
      model       = api_model,
      messages    = list(list(role = "user", content = prompt_text)),
      n           = n,
      temperature = temperature,
      max_tokens  = max_tokens
    )) |>
    req_retry(max_tries = 4, is_transient = \(resp) resp_status(resp) %in% c(429, 500, 503)) |>
    req_timeout(60)
}

# Pull the k completion texts out of one response (or character(0) on failure).
resp_texts <- function(resp) {
  if (!inherits(resp, "httr2_response")) return(character(0))
  body <- tryCatch(resp_body_json(resp), error = function(e) NULL)
  if (is.null(body$choices)) return(character(0))
  vapply(body$choices,
         function(ch) ch$message$content %||% NA_character_,
         character(1))
}

write_run_meta <- function(path, run_meta) {
  writeLines(jsonlite::toJSON(run_meta, auto_unbox = TRUE, pretty = TRUE), path)
}
```

## Run Pipeline

```{r run-pipeline}
if (!file.exists(norm_samples_path))
  stop("Missing: ", norm_samples_path, call. = FALSE)
if (!file.exists(instructions_rda_path))
  stop("Missing: ", instructions_rda_path, call. = FALSE)

api_key <- Sys.getenv(model_cfg$api_key_env)
if (!nzchar(api_key)) stop(model_cfg$api_key_env, " not set.", call. = FALSE)

responses_file <- "responses_sampled.csv"

norm_samples <- read.csv(norm_samples_path, stringsAsFactors = FALSE)
instructions <- read_norming_instructions(instructions_rda_path)
instr_map    <- build_instruction_map(instructions)
task_rows    <- build_tasks(instr_map, norm_samples,
                            include_constructs, exclude_constructs)
if (!nrow(task_rows)) stop("No tasks after filtering.", call. = FALSE)

prompts_df <- build_prompts(task_rows, norm_samples, instr_map,
                            max_words = 2000L,
                            words_cap = words_per_dim_cap, seed = sample_seed)

timestamp <- format(Sys.time(), "%Y%m%dT%H%M%S")
run_id    <- paste0(model_slug, "_sampled_", timestamp)
now_utc   <- format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
run_dir   <- ensure_dir(file.path(output_root, run_id))
tasks_dir <- ensure_dir(file.path(run_dir, "tasks"))

run_meta <- list(
  run_id = run_id, timestamp_utc = now_utc, model_slug = model_slug,
  api_model = model_cfg$api_model, provider = "openai",
  extraction = "sampled", client = "httr2",
  responses_file = responses_file, n_draws = n_draws, temperature = sample_temp,
  words_per_dim_cap = words_per_dim_cap %||% NA,
  n_prompts = nrow(prompts_df), tasks = sort(task_rows$task_id)
)
write_run_meta(file.path(run_dir, "run_meta.json"), run_meta)

message("Run ID    : ", run_id, "  (k=", n_draws, ", T=", sample_temp, ")")
message("Requests  : ", nrow(prompts_df), " (1 per word, n=", n_draws,
        " completions each)")

tasks <- sort(unique(prompts_df$task_id))
summary_rows <- list()
draw_rows    <- list()

for (j in seq_along(tasks)) {
  tid <- tasks[[j]]
  message("\n[", j, "/", length(tasks), "] ", tid)

  task_df <- filter(prompts_df, task_id == tid)
  s_min <- task_df$scale_min[[1]]; s_max <- task_df$scale_max[[1]]

  reqs <- lapply(paste0(task_df$instruction_text, "\n\nWord: ", task_df$word),
                 build_req, api_url = api_url, api_model = model_cfg$api_model,
                 api_key = api_key, n = n_draws, temperature = sample_temp,
                 max_tokens = max_tokens)

  resps <- req_perform_parallel(reqs, max_active = max_active,
                                on_error = "continue", progress = TRUE)

  draws_long <- bind_rows(lapply(seq_along(resps), function(i) {
    txts <- resp_texts(resps[[i]])
    if (!length(txts)) return(tibble(idx = i, draw = integer(0), rating = integer(0)))
    tibble(idx = i, draw = seq_along(txts),
           rating = vapply(txts, parse_integer_text, integer(1), s_min, s_max))
  })) |>
    mutate(word = task_df$word[idx], dimension = task_df$dimension[idx])

  task_summary <- draws_long |>
    group_by(word, dimension) |>
    summarise(
      n_valid        = sum(!is.na(rating)),
      expected       = mean(rating, na.rm = TRUE),       # MC expected value
      sampled_sd     = sd(rating, na.rm = TRUE),
      integer_rating = sample_mode(rating),              # sampled point estimate
      .groups = "drop"
    ) |>
    left_join(distinct(task_df, word, dimension, split, score),
              by = c("word", "dimension")) |>
    transmute(
      run_id = run_id, timestamp_utc = now_utc, model_slug = model_slug,
      model = model_cfg$api_model, word, dimension, split,
      human_score = score, extraction = "sampled",
      expected, integer_rating, coverage = n_valid / n_draws, sampled_sd
    )

  write.csv(task_summary, file.path(tasks_dir, paste0(tid, "_summary.csv")),
            row.names = FALSE)
  summary_rows[[j]] <- task_summary
  draw_rows[[j]] <- draws_long |>
    transmute(run_id = run_id, model_slug = model_slug, model = model_cfg$api_model,
              word, dimension, draw, rating)

  message("  done — ", nrow(task_summary), " words, ",
          round(mean(task_summary$coverage), 3), " mean coverage")
}

responses_out <- bind_rows(summary_rows)
write.csv(responses_out, file.path(run_dir, responses_file), row.names = FALSE)
write.csv(bind_rows(draw_rows), file.path(run_dir, "draws.csv"), row.names = FALSE)

message("\nWrote: ", file.path(run_dir, responses_file))
message("Wrote: ", file.path(run_dir, "draws.csv"))
message("Run complete: ", run_id)
```
