---
title: "Logprob Norming Efficacy — Paper 1 Analysis (multi-model horse race)"
always_allow_html: true
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 10
    fig_height: 6
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

```{r libraries}
library(dplyr)
library(readr)
library(tidyr)
library(jsonlite)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(ggplot2)
library(rlang)
```

```{r paths}
find_pkg_root <- function(start = getwd()) {
  cur <- normalizePath(start, mustWork = TRUE)
  repeat {
    if (file.exists(file.path(cur, "DESCRIPTION"))) return(cur)
    parent <- dirname(cur)
    if (identical(parent, cur)) break
    cur <- parent
  }
  stop("Could not find package root.", call. = FALSE)
}

base_dir <- find_pkg_root()
out_dir  <- file.path(base_dir, "analysis", "output")
runs_dir <- file.path(out_dir, "runs")

wide_path     <- file.path(out_dir, "core_vocab_wide.csv")
emb_pred_path <- file.path(out_dir, "embedding_model_predictions.csv")

# Embedding baselines included as methods in the horse race.
emb_methods <- c(glove2024 = "GloVe2024", bge_large = "BGE-large")

# Reference level for the abs_error mixed model (a strong supervised baseline).
horse_ref <- "BGE-large"

stopifnot(file.exists(wide_path), file.exists(emb_pred_path))

# prevalence_score is stored as a probit-transformed prevalence value derived
# from Pknown. The LLM was prompted on a 1-10 scale, so prevalence outputs are
# mapped back to Pknown and re-transformed onto the original score scale.
llm_prevalence_to_score <- function(x, dimension) {
  pknown <- pmin(pmax((x - 1) / 9, 0), 1)
  score  <- qnorm(0.005 + 0.99 * pknown)
  ifelse(dimension == "prevalence_score", score, x)
}
```

```{r discover-runs}
# A run is a model-prefixed folder carrying run_meta.json. Completeness is gated
# on the per-task file count: logprob runs emit a _summary + _distributions file
# per dimension (19 * 2 = 38); integer runs emit one file per dimension (19).
discover_run <- function(rd) {
  meta_path <- file.path(rd, "run_meta.json")
  if (!file.exists(meta_path)) return(NULL)            # excludes legacy runs
  m <- jsonlite::fromJSON(meta_path)
  n_task_files <- length(list.files(file.path(rd, "tasks")))
  need <- if (identical(m$extraction, "logprob")) 38L else 19L
  tibble(
    run_dir        = rd,
    run_id         = m$run_id,
    model_slug     = m$model_slug,
    provider       = m$provider %||% NA_character_,
    extraction     = m$extraction,
    responses_file = m$responses_file,
    n_task_files   = n_task_files,
    complete       = n_task_files == need
  )
}

run_dirs <- list.dirs(runs_dir, recursive = FALSE)
runs_tbl <- bind_rows(lapply(run_dirs, discover_run)) |>
  filter(complete) |>
  arrange(extraction, model_slug)

stopifnot(nrow(runs_tbl) > 0)

read_resp <- function(rd, rf) {
  p <- file.path(rd, rf)
  if (!file.exists(p)) {
    stop(sprintf("Run %s is tasks-complete but missing %s",
                 basename(rd), rf), call. = FALSE)
  }
  read_csv(p, show_col_types = FALSE)
}

# responses_*.csv already carries model_slug + extraction; only provider is new.
all_resp <- bind_rows(Map(read_resp, runs_tbl$run_dir, runs_tbl$responses_file)) |>
  left_join(runs_tbl |> select(run_id, provider), by = "run_id")
```

```{r dim-meta}
dim_groups <- tibble::tribble(
  ~dimension,                  ~group,       ~dim_label,
  "lancaster_auditory",        "Sensory",    "auditory",
  "lancaster_visual",          "Sensory",    "visual",
  "lancaster_haptic",          "Sensory",    "haptic",
  "lancaster_interoceptive",   "Sensory",    "interoceptive",
  "lancaster_gustatory",       "Sensory",    "gustatory",
  "lancaster_olfactory",       "Sensory",    "olfactory",
  "lancaster_foot_leg",        "Effector",   "foot/leg",
  "lancaster_hand_arm",        "Effector",   "hand/arm",
  "lancaster_head",            "Effector",   "head",
  "lancaster_mouth",           "Effector",   "mouth",
  "lancaster_torso",           "Effector",   "torso",
  "valence",                   "Affective",  "valence",
  "arousal",                   "Affective",  "arousal",
  "dominance",                 "Affective",  "dominance",
  "socialness",                "Lexical",    "socialness",
  "aoa",                       "Lexical",    "AoA",
  "concreteness",              "Lexical",    "concreteness",
  "iconicity",                 "Lexical",    "iconicity",
)
```

```{r load}
# Word-level covariates: computed once on the unique holdout words so the
# z-scores are not distorted by row duplication across models/dimensions.
covars <- read_csv(wide_path, show_col_types = FALSE) |>
  filter(split == "holdout") |>
  distinct(word, .keep_all = TRUE) |>
  mutate(
    log_n_synsets = log1p(wn_n_synsets),
    n_pos = as.integer((wn_n_noun > 0) + (wn_n_verb > 0) +
                         (wn_n_adj > 0) + (wn_n_adv > 0)),
    concreteness_z  = as.numeric(scale(concreteness)),
    freq_z          = as.numeric(scale(prevalence_freq_zipf)),
    log_n_synsets_z = as.numeric(scale(log_n_synsets)),
    n_pos_z         = as.numeric(scale(n_pos)),
    freq_band = factor(freq_band, levels = c(1, 2, 3),
                       labels = c("Low", "Mid", "High"))
  ) |>
  select(word, concreteness, prevalence_freq_zipf, wn_n_synsets,
         log_n_synsets, n_pos, concreteness_z, freq_z, log_n_synsets_z,
         n_pos_z, conc_q, freq_band)

# Long response frame across all included models. prevalence outputs mapped back
# to the score scale; prevalence_score dropped; restricted to holdout.
# The sampled run shares model_slug "gpt-4o-mini"; it is handled on its own in the
# Sampling section (read directly from its run dir) and must be excluded here so it
# does not double-count in the integer horse race or duplicate human_map rows.
dat <- all_resp |>
  filter(split == "holdout", dimension != "prevalence_score",
         extraction != "sampled") |>
  mutate(
    response_integer = llm_prevalence_to_score(integer_rating, dimension),
    expected         = llm_prevalence_to_score(expected, dimension)
  ) |>
  left_join(covars, by = "word")

dat_lp <- dat |> filter(extraction == "logprob", !is.na(expected))

# Canonical human-norm map (one value per word x dimension, model-invariant).
human_map <- dat |> distinct(word, dimension, human_score)

# Embedding-model predictions become additional methods in the horse race.
emb_long <- read_csv(emb_pred_path, show_col_types = FALSE) |>
  filter(embedding %in% names(emb_methods), split == "holdout") |>
  transmute(word, dimension,
            method    = unname(emb_methods[embedding]),
            predicted = predicted)

lexis_env <- new.env(parent = emptyenv())
load(file.path(base_dir, "data", "lexis_long.rda"), envir = lexis_env)
lexis_long <- lexis_env$lexis_long

lmer_ctrl <- lmerControl(optimizer = "bobyqa")

featured_slug <- grep("^gpt-4\\.1", unique(dat$model_slug), value = TRUE)[1]
mini_slug     <- grep("gpt-4o-mini", unique(dat_lp$model_slug), value = TRUE)[1]
stopifnot(length(featured_slug) == 1, !is.na(featured_slug),
          length(mini_slug) == 1, !is.na(mini_slug))
```

---

## Models Included

**What this section shows:** The runs admitted by the completeness gate (model-prefixed folder + full per-task file count). Logprob runs carry the probability-weighted `expected` value and a per-rating distribution; integer runs carry only the top-token rating. The integer rating is available for every model and drives the general horse race; the `expected` value and SD analyses are restricted to logprob models.

```{r models-table}
models_summary <- dat |>
  group_by(model_slug, provider, extraction) |>
  summarise(
    n_dimensions = n_distinct(dimension),
    n_words      = n_distinct(word),
    n_rows       = n(),
    .groups = "drop"
  ) |>
  arrange(extraction, model_slug)

knitr::kable(models_summary,
  caption = sprintf("Included runs (%d models; %d holdout dimensions)",
                    n_distinct(dat$model_slug), n_distinct(dat$dimension)))

logprob_models <- sort(unique(dat_lp$model_slug))
cat(sprintf("Logprob models (expected + SD eligible): %s\n",
            paste(logprob_models, collapse = ", ")))
```

---

## General Horse Race (all models, integer prediction)

**What this section shows:** Every model contributes its top-token integer rating, evaluated against the two supervised embedding baselines (GloVe300, BGE-large) on the same holdout set. This is a direct, head-to-head comparison of zero-shot LLM elicitation against distributional prediction. The descriptive r table shows where each method dominates; the mixed model formalizes mean accuracy differences after adjusting for word- and dimension-level difficulty.

```{r horse-data}
llm_int_long <- dat |>
  transmute(word, dimension,
            method    = paste0(model_slug, " (integer)"),
            predicted = response_integer)

horse_long <- bind_rows(llm_int_long, emb_long) |>
  inner_join(human_map, by = c("word", "dimension")) |>
  inner_join(covars,    by = "word") |>
  filter(!is.na(predicted)) |>
  mutate(abs_error = abs(human_score - predicted),
         method = factor(method))

if (horse_ref %in% levels(horse_long$method)) {
  horse_long$method <- relevel(horse_long$method, ref = horse_ref)
}
```

### Descriptive correlations

```{r horse-cors}
horse_cors <- horse_long |>
  group_by(dimension, method) |>
  summarise(r = cor(human_score, predicted, use = "complete.obs"),
            .groups = "drop")
```

#### Mean holdout r by method

```{r horse-mean-r}
horse_cors |>
  group_by(Method = method) |>
  summarise(`Mean r` = round(mean(r), 3), .groups = "drop") |>
  arrange(desc(`Mean r`)) |>
  knitr::kable(caption = sprintf("Mean holdout r across %d dimensions",
                                 n_distinct(horse_cors$dimension)))
```

#### Model wins

Highest r per dimension.

```{r horse-wins}
horse_cors |>
  group_by(dimension) |>
  slice_max(r, n = 1, with_ties = FALSE) |>
  ungroup() |>
  count(Method = method, name = "Wins") |>
  arrange(desc(Wins)) |>
  knitr::kable()
```

#### Per-dimension r table

```{r horse-cors-table}
horse_cors |>
  left_join(dim_groups, by = "dimension") |>
  mutate(r = round(r, 3)) |>
  select(Group = group, Dimension = dim_label, method, r) |>
  pivot_wider(names_from = method, values_from = r) |>
  arrange(match(Group, c("Sensory", "Effector", "Affective", "Lexical"))) |>
  knitr::kable(caption = "Holdout Pearson r by dimension and method")
```

### Mixed model: method effect on absolute prediction error

**What this shows:** Absolute prediction error as the outcome, method as a fixed factor with one level per model plus each embedding baseline, and crossed random effects for word and dimension. Dimension random effects absorb that some constructs are harder for all methods; word random effects absorb consistently hard items. Reference level: `r horse_ref`.

```{r horse-lmer}
m_horse <- lmer(
  abs_error ~ method + concreteness_z + freq_z + log_n_synsets_z + n_pos_z +
    (1 | word) + (1 | dimension),
  data = horse_long, REML = TRUE, control = lmer_ctrl
)

print(round(summary(m_horse)$coefficients, 4))
```

```{r horse-lmer-variances}
print(VarCorr(m_horse))
```

#### Pairwise contrasts vs reference

Negative estimate = lower abs_error than the reference method.

```{r horse-contrasts}
coefs <- fixef(m_horse)
ses   <- sqrt(diag(vcov(m_horse)))
method_terms <- grep("^method", names(coefs), value = TRUE)

tibble(
  method   = sub("^method", "", method_terms),
  estimate = round(coefs[method_terms], 4),
  se       = round(ses[method_terms], 4),
  t        = round(coefs[method_terms] / ses[method_terms], 4),
  p        = round(2 * pt(abs(coefs[method_terms] / ses[method_terms]),
                          df = df.residual(m_horse), lower.tail = FALSE), 5)
) |>
  mutate(sig = dplyr::case_when(
    p < 0.001 ~ "***", p < 0.01 ~ "**", p < 0.05 ~ "*", TRUE ~ "")) |>
  arrange(estimate) |>
  print()
```

### Collinearity: n_pos vs log synset count

**What this shows:** The two polysemy operationalizations — fine-grained sense
multiplicity (`log_n_synsets`) and grammatical versatility (`n_pos`, range 1–4) — enter
the moderation models jointly, so each is estimated as a partial effect. Before
interpreting those partial effects we check that the two are sufficiently independent.
Positive correlation is expected (more POS categories → more synsets), but `n_pos` is
coarse, so the variance inflation factors should remain well below conventional concern
(< 5). Source for the manuscript's "Pearson *r* = .56 … VIF 1.5 and 1.8" (¶ on the two
polysemy measures). Computed on the unique holdout words (`covars`).

```{r polysemy-collinearity}
r_poly <- cor(covars$log_n_synsets, covars$n_pos, use = "complete.obs")
cat(sprintf("Pearson r(log_n_synsets, n_pos) = %.3f\n", r_poly))

vif_tbl <- tibble(
  term = c("log_n_synsets_z", "n_pos_z"),
  VIF  = c(
    1 / (1 - summary(lm(log_n_synsets_z ~ freq_z + n_pos_z,
                        data = covars))$r.squared),
    1 / (1 - summary(lm(n_pos_z ~ freq_z + log_n_synsets_z,
                        data = covars))$r.squared)
  )
) |>
  mutate(VIF = round(VIF, 2))

knitr::kable(vif_tbl,
  caption = "Variance inflation factors (other covariates partialled out)")
```

### Method × covariate interactions

**What this shows:** Whether each method's error advantage is uniform across the lexical space or concentrated in particular regions. Each LRT adds one interaction block (concreteness, frequency, or the polysemy pair `log_n_synsets_z` + `n_pos_z`) to the main-effects model. A significant test means the method ordering depends on that lexical property.

```{r horse-interactions}
m_cov_main <- lmer(
  abs_error ~ method + concreteness_z + freq_z + log_n_synsets_z + n_pos_z +
    (1 | word) + (1 | dimension),
  data = horse_long, REML = FALSE, control = lmer_ctrl
)

m_conc_int <- lmer(
  abs_error ~ method * concreteness_z + freq_z + log_n_synsets_z + n_pos_z +
    (1 | word) + (1 | dimension),
  data = horse_long, REML = FALSE, control = lmer_ctrl
)

m_freq_int <- lmer(
  abs_error ~ method + concreteness_z + method * freq_z +
    log_n_synsets_z + n_pos_z +
    (1 | word) + (1 | dimension),
  data = horse_long, REML = FALSE, control = lmer_ctrl
)

m_poly_int <- lmer(
  abs_error ~ method + concreteness_z + freq_z +
    method * (log_n_synsets_z + n_pos_z) +
    (1 | word) + (1 | dimension),
  data = horse_long, REML = FALSE, control = lmer_ctrl
)

m_interact_ml <- lmer(
  abs_error ~ method * (concreteness_z + freq_z + log_n_synsets_z + n_pos_z) +
    (1 | word) + (1 | dimension),
  data = horse_long, REML = FALSE, control = lmer_ctrl
)
```

#### LRT: method × concreteness

```{r horse-lrt-conc}
print(anova(m_cov_main, m_conc_int))
```

#### LRT: method × frequency

```{r horse-lrt-freq}
print(anova(m_cov_main, m_freq_int))
```

#### LRT: method × polysemy (synsets + n_pos jointly)

```{r horse-lrt-poly}
print(anova(m_cov_main, m_poly_int))
```

#### LRT: all interaction blocks vs mains only

```{r horse-lrt-all}
print(anova(m_cov_main, m_interact_ml))
```

---

## Part A: Method × Lexical Moderators (range-normalized error)

**What this shows:** The Part A moderation model behind the representation benchmark. Method is a two-level factor — the strongest LLM in the integer horse race (`gpt-4.1`) and BGE-large. Because the dimensions use different response scales (0–5, 1–9, 1–25), the outcome is **range-normalized absolute error** (absolute error ÷ the dimension's response-scale span), so dimensions contribute on a common 0–1 footing. All four lexical covariates are interacted with method jointly; an omnibus LRT plus a concreteness-added-last LRT establish whether the embedding advantage depends on lexical properties. Expected-value-specific lexical uniformity is tested later in Part B.

```{r moderation-data}
scale_span <- c(
  lancaster_auditory = 5, lancaster_visual = 5, lancaster_haptic = 5,
  lancaster_interoceptive = 5, lancaster_gustatory = 5, lancaster_olfactory = 5,
  lancaster_foot_leg = 5, lancaster_hand_arm = 5, lancaster_head = 5,
  lancaster_mouth = 5, lancaster_torso = 5, valence = 8, arousal = 8,
  dominance = 8, socialness = 6, aoa = 24, concreteness = 4, iconicity = 6)

featured_int <- dat |>
  filter(model_slug == featured_slug) |>
  transmute(word, dimension, method = paste0(featured_slug, " (integer)"),
            predicted = response_integer, human_score)

mod_dat <- bind_rows(
  featured_int,
  emb_long |> filter(method == "BGE-large") |>
    inner_join(human_map, by = c("word", "dimension"))
) |>
  inner_join(covars, by = "word") |>
  filter(!is.na(predicted)) |>
  mutate(err = abs(human_score - predicted) / scale_span[dimension],
         method = relevel(factor(method), ref = paste0(featured_slug, " (integer)")))
```

### Method main effects (range-normalized)

```{r moderation-main}
m_main <- lmer(
  err ~ method + concreteness_z + freq_z + log_n_synsets_z + n_pos_z +
    (1 | word) + (1 | dimension),
  data = mod_dat, REML = TRUE, control = lmer_ctrl)
print(round(summary(m_main)$coefficients, 4))
print(VarCorr(m_main))
```

### Full joint interaction model + omnibus LRTs

```{r moderation-joint}
m_joint <- lmer(
  err ~ method * (concreteness_z + freq_z + log_n_synsets_z + n_pos_z) +
    (1 | word) + (1 | dimension),
  data = mod_dat, REML = TRUE, control = lmer_ctrl)
print(round(summary(m_joint)$coefficients, 4))

# Omnibus: all method x covariate interactions; and concreteness added last
m_main_ml <- lmer(
  err ~ method + concreteness_z + freq_z + log_n_synsets_z + n_pos_z +
    (1 | word) + (1 | dimension),
  data = mod_dat, REML = FALSE, control = lmer_ctrl)
m_joint_ml <- lmer(
  err ~ method * (concreteness_z + freq_z + log_n_synsets_z + n_pos_z) +
    (1 | word) + (1 | dimension),
  data = mod_dat, REML = FALSE, control = lmer_ctrl)
m_noconc_ml <- lmer(
  err ~ method * (freq_z + log_n_synsets_z + n_pos_z) + concreteness_z +
    (1 | word) + (1 | dimension),
  data = mod_dat, REML = FALSE, control = lmer_ctrl)
cat("Omnibus (all interactions):\n");        print(anova(m_main_ml, m_joint_ml))
cat("\nConcreteness x method (added last):\n"); print(anova(m_noconc_ml, m_joint_ml))
```

### Residualized efficacy (3 methods)

```{r moderation-resid}
resid_cor_vec <- function(human, pred, conc, freq, log_syn, npos) {
  df <- data.frame(human, pred, conc, freq, log_syn, npos)
  df <- df[complete.cases(df), ]
  if (nrow(df) < 10) return(NA_real_)
  cor(residuals(lm(human ~ conc + freq + log_syn + npos, df)),
      residuals(lm(pred  ~ conc + freq + log_syn + npos, df)))
}
mod_dat |>
  group_by(method, dimension) |>
  group_modify(~ tibble(r_res = resid_cor_vec(.x$human_score, .x$predicted,
    .x$concreteness_z, .x$freq_z, .x$log_n_synsets_z, .x$n_pos_z))) |>
  group_by(Method = method) |>
  summarise(`Residual r` = round(mean(r_res, na.rm = TRUE), 3), .groups = "drop") |>
  arrange(desc(`Residual r`)) |>
  knitr::kable(caption = "Residualized mean r by method (four lexical covariates)")
```

---

## Part A: Norm Reliability Ceiling

**What this shows:** Whether the LLM-vs-embedding win/loss pattern is an artifact of differential human-norm reliability (review concern: noisier norms cap achievable *r* for every method). We compute the reliability of each dimension's *published mean* norm from the shipped per-item SD and rater count — true-score variance ÷ observed between-item variance, where each item's measurement-error variance is `sd^2 / n_ratings`. The Lancaster norms carry per-item SD but no rating count, so `n_ratings` for the six perceptual modalities is taken as `N_known.perceptual` and for the five action modalities as `N_known.action` from the source file (each participant who knew a word rated it on all perceptual or all action dimensions). The test is whether the per-dimension embedding advantage (BGE − gpt-4.1 integer) tracks reliability; the key counter-evidence is whether high-reliability sensorimotor norms still show large embedding advantages.

```{r reliability}
analyzed_dims <- sort(unique(dat_lp$dimension))

perc_dims <- c("lancaster_auditory","lancaster_gustatory","lancaster_haptic",
               "lancaster_interoceptive","lancaster_olfactory","lancaster_visual")
actn_dims <- c("lancaster_foot_leg","lancaster_hand_arm","lancaster_head",
               "lancaster_mouth","lancaster_torso")

lanc_n <- read_csv(
  file.path(base_dir, "datasets", "lancaster",
            "Lancaster_sensorimotor_norms_for_39707_words.csv"),
  show_col_types = FALSE) |>
  transmute(word = tolower(Word),
            N_perceptual = `N_known.perceptual`,
            N_action     = `N_known.action`)

holdout_words <- covars$word

rel_long <- lexis_long |>
  filter(dimension %in% analyzed_dims, word %in% holdout_words,
         !is.na(sd), !is.na(mean)) |>
  left_join(lanc_n, by = "word") |>
  mutate(n_ratings = case_when(
    dimension %in% perc_dims ~ as.numeric(N_perceptual),
    dimension %in% actn_dims ~ as.numeric(N_action),
    TRUE                     ~ as.numeric(n_ratings))) |>
  filter(!is.na(n_ratings), n_ratings > 1) |>
  group_by(word, dimension) |>            # collapse cross-dataset dupes
  slice_max(n_ratings, n = 1, with_ties = FALSE) |>
  ungroup()

reliability_tbl <- rel_long |>
  group_by(dimension) |>
  summarise(n_items     = n(),
            med_nrater  = median(n_ratings),
            var_means   = var(mean),
            mean_errvar = mean(sd^2 / n_ratings),
            .groups = "drop") |>
  mutate(reliability = pmax(0, (var_means - mean_errvar) / var_means))
```

### Reliability with per-dimension benchmark r

```{r reliability-table}
r_llm_int <- dat |>
  filter(model_slug == featured_slug) |>
  group_by(dimension) |>
  summarise(r_llm_int = cor(human_score, response_integer, use = "complete.obs"),
            .groups = "drop")
r_bge <- horse_cors |>
  filter(method == "BGE-large") |>
  transmute(dimension, r_bge = r)

rel_bench <- reliability_tbl |>
  left_join(r_llm_int, by = "dimension") |>
  left_join(r_bge,     by = "dimension") |>
  left_join(dim_groups, by = "dimension") |>
  mutate(advantage = r_bge - r_llm_int) |>
  arrange(desc(reliability))

rel_bench |>
  transmute(Group = group, Dimension = dim_label,
            `N raters` = med_nrater,
            Reliability = round(reliability, 3),
            `r LLM(int)` = round(r_llm_int, 3),
            `r BGE` = round(r_bge, 3),
            `BGE adv.` = round(advantage, 3)) |>
  knitr::kable(caption = "Per-dimension reliability of the published mean norm, with benchmark r")
```

### Does the embedding advantage track reliability?

```{r reliability-test}
ct <- cor.test(rel_bench$advantage, rel_bench$reliability)
cat(sprintf("cor(BGE advantage, reliability) = %.3f, p = %.3f, n = %d\n",
            ct$estimate, ct$p.value, nrow(rel_bench)))
cat(sprintf("cor(r LLM(int), reliability)     = %.3f\n",
            cor(rel_bench$r_llm_int, rel_bench$reliability)))
cat(sprintf("cor(r BGE, reliability)          = %.3f\n",
            cor(rel_bench$r_bge, rel_bench$reliability)))

rel_bench <- rel_bench |> mutate(llm_wins = r_llm_int > r_bge)
cat(sprintf("\nmean reliability where LLM wins = %.3f; where BGE wins = %.3f\n",
            mean(rel_bench$reliability[rel_bench$llm_wins]),
            mean(rel_bench$reliability[!rel_bench$llm_wins])))

# Counterexample: high-reliability sensorimotor norms with large embedding advantage
cat("\nHigh-reliability (>= .80) Lancaster norms with BGE advantage >= .10:\n")
rel_bench |>
  filter(group %in% c("Sensory","Effector"), reliability >= .80, advantage >= .10) |>
  transmute(dim_label, reliability = round(reliability,3),
            r_llm_int = round(r_llm_int,3), r_bge = round(r_bge,3),
            advantage = round(advantage,3)) |>
  as.data.frame() |> print(row.names = FALSE)
```

### Attenuation-corrected r (sensitivity)

```{r reliability-attenuation}
rel_bench |>
  mutate(`r LLM(int)` = round(r_llm_int, 3),
         `r LLM corr` = round(pmin(1, r_llm_int / sqrt(reliability)), 3),
         `r BGE` = round(r_bge, 3),
         `r BGE corr` = round(pmin(1, r_bge / sqrt(reliability)), 3)) |>
  transmute(Dimension = dim_label, Reliability = round(reliability, 3),
            `r LLM(int)`, `r LLM corr`, `r BGE`, `r BGE corr`) |>
  knitr::kable(caption = "Attenuation-corrected holdout r (r / sqrt(reliability))")
```

---

## Part A: Per-Dimension Bootstrap Confidence Intervals (all 18 dimensions)

**What this shows:** Uniform 95% bootstrap CIs on the holdout *r* for the Part A
division-of-labor comparison — gpt-4.1 integer vs BGE-large — across **all 18
dimensions**, so the per-dimension claims rest on intervals rather than point estimates.
A paired bootstrap (the same resampled words score both methods) makes the `BGE - LLM(int)` advantage CI correct; a CI excluding 0 marks a
reliable per-dimension winner, and an overlapping CI marks a tentative one. The seed is
fixed and reported in the manuscript's Data & Code Availability.

```{r per-dim-bootstrap}
set.seed(20260604)
n_boot <- 2000

boot_dat <- dat |>
  filter(model_slug == featured_slug) |>
  transmute(word, dimension, human_score, llm_int = response_integer) |>
  inner_join(
    emb_long |> filter(method == "BGE-large") |>
      transmute(word, dimension, bge = predicted),
    by = c("word", "dimension")) |>
  filter(!is.na(human_score), !is.na(llm_int), !is.na(bge))

boot_one_dim <- function(df) {
  n   <- nrow(df)
  idx <- matrix(sample.int(n, n * n_boot, replace = TRUE), nrow = n)
  bl  <- apply(idx, 2, function(i) cor(df$human_score[i], df$llm_int[i]))
  bb  <- apply(idx, 2, function(i) cor(df$human_score[i], df$bge[i]))
  est_llm <- cor(df$human_score, df$llm_int)
  est_bge <- cor(df$human_score, df$bge)
  qa  <- quantile(bb - bl, c(.025, .975))
  tibble(
    n_words   = n,
    r_llm_int = est_llm,
    llm_lo = quantile(bl, .025), llm_hi = quantile(bl, .975),
    r_bge  = est_bge,
    bge_lo = quantile(bb, .025), bge_hi = quantile(bb, .975),
    advantage = est_bge - est_llm,
    adv_lo = qa[[1]], adv_hi = qa[[2]])
}

boot_ci <- boot_dat |>
  group_by(dimension) |>
  group_modify(~ boot_one_dim(.x)) |>
  ungroup() |>
  left_join(dim_groups, by = "dimension") |>
  arrange(group, desc(r_llm_int))

fmt_ci <- function(est, lo, hi) sprintf("%.3f [%.3f, %.3f]", est, lo, hi)

boot_ci |>
  transmute(
    Group = group, Dimension = dim_label, `N words` = n_words,
    `LLM(int) r [95% CI]` = fmt_ci(r_llm_int, llm_lo, llm_hi),
    `BGE r [95% CI]`      = fmt_ci(r_bge, bge_lo, bge_hi),
    `BGE adv. [95% CI]`   = fmt_ci(advantage, adv_lo, adv_hi)) |>
  knitr::kable(
    caption = paste0(
      "Per-dimension holdout r with 95% bootstrap CIs (", n_boot,
      " resamples, seed 20260604): gpt-4.1 integer vs BGE-large. ",
      "'BGE adv.' is the paired BGE - LLM(int) difference; a CI excluding 0 ",
      "indicates a reliable per-dimension winner."))
```

---

## Part A: Predicted-Range Calibration (SD ratio)

**What this shows:** Whether a method's predictions occupy the same dynamic range as the human norms. SD ratio = SD(predictions) ⁄ SD(human) per dimension. Ridge regression estimates a conditional mean, so its fitted values are attenuated by ≈ the validation correlation (regression to the mean) → SD ratio < 1. Direct LLM elicitation is not a least-squares estimator and carries no such attenuation; the empirical question is whether its spread is otherwise calibrated.

```{r sd-ratio}
emb_ratio <- emb_long |>
  inner_join(human_map, by = c("word", "dimension")) |>
  group_by(method, dimension) |>
  summarise(sd_ratio = sd(predicted) / sd(human_score),
            r = cor(human_score, predicted), .groups = "drop")

llm_ratio <- dat |>
  filter(model_slug == featured_slug) |>
  group_by(dimension) |>
  summarise(`gpt-4.1 (integer)` = sd(response_integer) / sd(human_score),
            .groups = "drop")

cat("Mean SD ratio across 18 dimensions:\n")
emb_ratio |> group_by(method) |>
  summarise(`SD ratio` = round(mean(sd_ratio), 3),
            `mean r` = round(mean(r), 3), .groups = "drop") |>
  knitr::kable() |> print()
cat(sprintf("gpt-4.1 integer: %.3f\n",
            mean(llm_ratio$`gpt-4.1 (integer)`)))

cat("\nLLM integer SD ratio by construct group:\n")
llm_ratio |> left_join(dim_groups, by = "dimension") |>
  group_by(group) |>
  summarise(sd_ratio = round(mean(`gpt-4.1 (integer)`), 2), .groups = "drop") |>
  as.data.frame() |> print(row.names = FALSE)
```

---


## Part A/B Bridge: Human Score Distributions

**What this shows:** The manuscript invokes floor-heavy dimensions and scale compression when interpreting the division of labor and the SD channel. This figure visualizes the human holdout score distributions by dimension, making clear which norms are broad and graded and which are sparse/floor-bound.

```{r human-score-distributions, fig.width=10, fig.height=7}
human_dist <- human_map |>
  left_join(dim_groups, by = "dimension") |>
  mutate(dim_label = factor(dim_label, levels = dim_groups$dim_label),
         group = factor(group, levels = c("Sensory", "Effector", "Affective", "Lexical")))

floor_tbl <- human_dist |>
  group_by(group, dim_label) |>
  summarise(
    n = n(),
    min_score = min(human_score, na.rm = TRUE),
    pct_at_floor = mean(human_score == min_score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(pct_at_floor))

floor_tbl |>
  mutate(`% at floor` = round(100 * pct_at_floor, 1),
         min_score = round(min_score, 3)) |>
  select(Group = group, Dimension = dim_label, N = n,
         Floor = min_score, `% at floor`) |>
  knitr::kable(caption = "Human holdout score floor mass by dimension")

p_dist <- ggplot(human_dist, aes(human_score)) +
  geom_histogram(bins = 25, fill = "#4C78A8", color = "white", linewidth = 0.15) +
  facet_wrap(~ dim_label, scales = "free", ncol = 6) +
  labs(x = "Human norm score", y = "Holdout words",
       title = "Human norm score distributions by dimension") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())

ggsave("fig_human_score_distributions.png", p_dist, width = 10, height = 7, dpi = 150)
p_dist
```

## Part B: Expected Value vs Integer Rating (logprob models)

**What this shows:** The primary extraction-rule hypothesis. Human aggregate norms are expected values; an LLM's top-sampled integer is a point estimate that discards distributional information. The logprob-weighted expected value (Σ rating × p(rating)) is structurally analogous to the human mean. This test asks whether aligning the mathematical object improves agreement with human norms — first per logprob model, then pooled across models.

### Per-model paired test

Paired t-test on Fisher-z transformed correlations; each dimension contributes one paired observation (expected vs integer) within a model.

```{r ei-cors}
ei_cors <- dat_lp |>
  group_by(model_slug, dimension) |>
  summarise(
    r_expected = cor(human_score, expected,         use = "complete.obs"),
    r_integer  = cor(human_score, response_integer, use = "complete.obs"),
    mean_cov   = mean(coverage, na.rm = TRUE),
    .groups    = "drop"
  )

ei_test <- ei_cors |>
  group_by(model_slug) |>
  group_modify(~ {
    tt <- t.test(atanh(.x$r_expected), atanh(.x$r_integer), paired = TRUE)
    tibble(
      n_dim         = nrow(.x),
      mean_r_exp    = round(mean(.x$r_expected), 3),
      mean_r_int    = round(mean(.x$r_integer), 3),
      mean_delta    = round(mean(.x$r_expected - .x$r_integer), 3),
      t             = round(unname(tt$statistic), 3),
      df            = unname(tt$parameter),
      p             = round(tt$p.value, 5),
      n_exp_gt_int  = sum(.x$r_expected > .x$r_integer)
    )
  }) |>
  ungroup()

knitr::kable(ei_test, caption = "Expected vs integer, per logprob model")
```

#### Per-model per-dimension detail

```{r ei-detail}
ei_cors |>
  left_join(dim_groups, by = "dimension") |>
  mutate(delta = r_expected - r_integer,
         across(c(r_expected, r_integer, delta, mean_cov), ~ round(.x, 3))) |>
  select(Model = model_slug, Dimension = dim_label, Group = group,
         `r (expected)` = r_expected, `r (integer)` = r_integer,
         Delta = delta, Coverage = mean_cov) |>
  arrange(Model, desc(Delta)) |>
  knitr::kable()
```

### Pooled test across models

**What this shows:** A single mixed model over the stacked logprob models with extraction (integer vs expected) as a within factor and crossed random effects for word, dimension, and model. A significant negative coefficient on `expected` means the expected value lowers absolute error relative to the integer rating, generalizing across models.

```{r ei-pooled}
ei_long <- dat_lp |>
  select(word, dimension, model_slug, human_score,
         integer = response_integer, expected) |>
  pivot_longer(c(integer, expected),
               names_to = "extraction", values_to = "predicted") |>
  filter(!is.na(predicted)) |>
  mutate(extraction = factor(extraction, levels = c("integer", "expected")),
         abs_error  = abs(human_score - predicted))

m_ei <- lmer(
  abs_error ~ extraction +
    (1 | word) + (1 | dimension) + (1 | model_slug),
  data = ei_long, REML = TRUE, control = lmer_ctrl
)

print(round(summary(m_ei)$coefficients, 4))
print(VarCorr(m_ei))
```


### EV lexical uniformity

**What this shows:** Whether the EV improvement is itself moderated by lexical properties. This is separate from the Part A BGE-vs-LLM moderator model: here the only contrast is extraction within the logprob-capable LLMs (integer vs expected), tested against the same covariates. Non-significant extraction × covariate terms support the claim that EV is a measurement refinement rather than a lexical-region-specific representation gain.

```{r ev-uniformity}
ev_uniform_dat <- dat_lp |>
  select(word, dimension, model_slug, human_score,
         integer = response_integer, expected) |>
  pivot_longer(c(integer, expected),
               names_to = "extraction", values_to = "predicted") |>
  inner_join(covars, by = "word") |>
  filter(!is.na(predicted)) |>
  mutate(extraction = relevel(factor(extraction), ref = "integer"),
         err = abs(human_score - predicted) / scale_span[dimension])

m_ev_uniform <- lmer(
  err ~ extraction * (concreteness_z + freq_z + log_n_synsets_z + n_pos_z) +
    (1 | word) + (1 | dimension) + (1 | model_slug),
  data = ev_uniform_dat, REML = TRUE, control = lmer_ctrl
)
print(round(summary(m_ev_uniform)$coefficients, 4))
```

### EV-credited horse race

**What this shows:** Whether crediting the two OpenAI logprob models with their expected values changes the representation benchmark. Non-logprob LLMs remain integer-only; embeddings are unchanged. If the EV changes only a small number of per-dimension winners, the EV is a measurement gain rather than a representation-level reversal.

```{r ev-credited-horse-race}
ev_llm_long <- dat |>
  mutate(pred_ev_credit = if_else(extraction == "logprob" & !is.na(expected),
                                  expected, response_integer)) |>
  transmute(word, dimension,
            method = paste0(model_slug, if_else(extraction == "logprob",
                                                " (expected)", " (integer)")),
            predicted = pred_ev_credit)

ev_horse_long <- bind_rows(ev_llm_long, emb_long) |>
  inner_join(human_map, by = c("word", "dimension")) |>
  filter(!is.na(predicted))

ev_horse_cors <- ev_horse_long |>
  group_by(dimension, method) |>
  summarise(r = cor(human_score, predicted, use = "complete.obs"),
            .groups = "drop")

cat("EV-credited mean r by method:\n")
ev_horse_cors |>
  group_by(Method = method) |>
  summarise(`Mean r` = round(mean(r), 3), .groups = "drop") |>
  arrange(desc(`Mean r`)) |>
  knitr::kable() |> print()

cat("\nEV-credited wins by method:\n")
ev_winners <- ev_horse_cors |>
  group_by(dimension) |>
  slice_max(r, n = 1, with_ties = FALSE) |>
  ungroup()
ev_winners |>
  count(Method = method, name = "Wins") |>
  arrange(desc(Wins)) |>
  knitr::kable() |> print()

cat("\nDimensions whose winner changes after EV crediting:\n")
int_winners <- horse_cors |>
  group_by(dimension) |>
  slice_max(r, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(dimension, integer_winner = as.character(method), integer_r = r)
ev_winners |>
  transmute(dimension, ev_winner = as.character(method), ev_r = r) |>
  inner_join(int_winners, by = "dimension") |>
  filter(ev_winner != integer_winner) |>
  left_join(dim_groups, by = "dimension") |>
  select(Group = group, Dimension = dim_label,
         integer_winner, integer_r, ev_winner, ev_r) |>
  mutate(across(c(integer_r, ev_r), ~ round(.x, 3))) |>
  knitr::kable() |> print()

cat("\nEV-credited per-dimension r table:\n")
ev_horse_cors |>
  left_join(dim_groups, by = "dimension") |>
  mutate(r = round(r, 3)) |>
  select(Group = group, Dimension = dim_label, method, r) |>
  pivot_wider(names_from = method, values_from = r) |>
  arrange(match(Group, c("Sensory", "Effector", "Affective", "Lexical"))) |>
  knitr::kable() |> print()
```

### Resolution check: rounding the continuous target and the EV

**What this shows:** Why EV gains are small but systematic. Published human norms are continuous means; integer ratings cannot occupy the sub-integer band of the target. Rounding the human norm estimates how much target signal lives between integers. Rounding the LLM expected value tests whether the EV gain is mostly sub-integer resolution.

```{r ev-resolution-rounding}
round_dat <- dat_lp |>
  filter(model_slug == mini_slug) |>
  select(word, dimension, human_score, response_integer, expected) |>
  filter(!is.na(human_score), !is.na(response_integer), !is.na(expected))

round_summary <- tibble(
  check = c(
    "cor(human, round(human))",
    "cor(human, LLM integer)",
    "cor(human, LLM expected)",
    "cor(human, round(LLM expected))"
  ),
  r = c(
    cor(round_dat$human_score, round(round_dat$human_score)),
    cor(round_dat$human_score, round_dat$response_integer),
    cor(round_dat$human_score, round_dat$expected),
    cor(round_dat$human_score, round(round_dat$expected))
  )
)

r_int <- round_summary$r[round_summary$check == "cor(human, LLM integer)"]
r_exp <- round_summary$r[round_summary$check == "cor(human, LLM expected)"]
r_exp_round <- round_summary$r[round_summary$check == "cor(human, round(LLM expected))"]

round_summary |>
  mutate(r = round(r, 3)) |>
  knitr::kable(caption = "Resolution check on the full holdout (gpt-4o-mini)")

cat(sprintf("EV gain erased by rounding EV: %.1f%%\n",
            100 * (r_exp - r_exp_round) / (r_exp - r_int)))
cat(sprintf("Cells where rounded EV differs from integer response: %.1f%%\n",
            100 * mean(round(round_dat$expected) != round_dat$response_integer)))
```

---

## Part B: Sampling — Recovering the Expected Value Without Logprobs

**What this shows:** The logprob expected value above needs an API that exposes token
log-probabilities. Here we test whether the *same* quantity is recoverable by sampling
alone — averaging temperature-1.0 draws — on the sampled gpt-4o-mini run. If the sampled
mean and the logprob expected value are two estimators of one integral over the model's
output distribution, they should agree, and the sampled route should match human norms
just as well. The sampled run (`gpt-4o-mini_sampled_…`) is auto-discovered; all numbers
below are computed here rather than read from a side file.

```{r sampling-load}
sampled_dir <- list.dirs(runs_dir, recursive = FALSE)
sampled_dir <- sampled_dir[grepl("gpt-4o-mini_sampled", basename(sampled_dir))]
sampled_dir <- sort(sampled_dir) |> tail(1)
stopifnot(length(sampled_dir) == 1, dir.exists(sampled_dir))

# Per-item sampled expected value (mean of the 20 draws), holdout, 18 analyzed dims.
sampled_ev <- read_csv(file.path(sampled_dir, "responses_sampled.csv"),
                       show_col_types = FALSE) |>
  filter(split == "holdout", dimension %in% dim_groups$dimension) |>
  transmute(word, dimension, sampled_exp = expected)

# Logprob expected value from the same model (gpt-4o-mini), same holdout x dims.
logprob_ev <- dat_lp |>
  filter(model_slug == mini_slug) |>
  transmute(word, dimension, logprob_exp = expected)

ev_join <- sampled_ev |>
  inner_join(logprob_ev, by = c("word", "dimension")) |>
  inner_join(human_map,  by = c("word", "dimension")) |>
  filter(!is.na(sampled_exp), !is.na(logprob_exp), !is.na(human_score))

cat(sprintf("n cells = %d across %d dimensions\n",
            nrow(ev_join), n_distinct(ev_join$dimension)))
```

### Equivalence of the two estimators

```{r sampling-equivalence}
cat(sprintf("r(sampled EV, logprob EV)  = %.3f\n",
            cor(ev_join$sampled_exp, ev_join$logprob_exp)))
cat(sprintf("r(sampled EV, human norm)  = %.3f\n",
            cor(ev_join$sampled_exp, ev_join$human_score)))
cat(sprintf("r(logprob EV, human norm)  = %.3f\n",
            cor(ev_join$logprob_exp, ev_join$human_score)))

ev_join |>
  group_by(dimension) |>
  summarise(r_sl = cor(sampled_exp, logprob_exp), .groups = "drop") |>
  left_join(dim_groups, by = "dimension") |>
  summarise(min_r = min(r_sl), min_dim = dim_label[which.min(r_sl)],
            max_r = max(r_sl), max_dim = dim_label[which.max(r_sl)]) |>
  print()
```

### How many draws are needed (k-sweep)

```{r sampling-ksweep}
set.seed(20260604)
k_grid   <- c(1, 2, 3, 5, 10, 15, 20)
n_resamp <- 200

draws <- read_csv(file.path(sampled_dir, "draws.csv"), show_col_types = FALSE) |>
  filter(dimension %in% dim_groups$dimension)

# Wide matrix: one row per (word, dimension), columns = the 20 draws; aligned to
# the logprob EV and the human norm.
draws_wide <- draws |>
  select(word, dimension, draw, rating) |>
  pivot_wider(names_from = draw, values_from = rating, names_prefix = "d") |>
  inner_join(logprob_ev, by = c("word", "dimension")) |>
  inner_join(human_map,  by = c("word", "dimension"))

draw_cols <- grep("^d[0-9]+$", names(draws_wide), value = TRUE)
D    <- as.matrix(draws_wide[, draw_cols])   # n_items x 20
lp   <- draws_wide$logprob_exp
hu   <- draws_wide$human_score
n_dr <- ncol(D)

ksweep <- purrr::map_dfr(k_grid, function(k) {
  rl <- numeric(n_resamp); rh <- numeric(n_resamp)
  for (b in seq_len(n_resamp)) {
    cols  <- sample.int(n_dr, k)             # k distinct draws of the 20
    m     <- rowMeans(D[, cols, drop = FALSE], na.rm = TRUE)
    rl[b] <- cor(m, lp, use = "complete.obs")
    rh[b] <- cor(m, hu, use = "complete.obs")
  }
  tibble(k = k,
         r_logprob = mean(rl), rlp_lo = quantile(rl, .025), rlp_hi = quantile(rl, .975),
         r_human   = mean(rh), rh_lo  = quantile(rh, .025), rh_hi  = quantile(rh, .975))
})

ceiling_h <- cor(lp, hu)

ksweep |>
  transmute(`Draws (k)` = k,
            `r vs logprob EV`  = sprintf("%.3f [%.3f, %.3f]", r_logprob, rlp_lo, rlp_hi),
            `r vs human norms` = sprintf("%.3f [%.3f, %.3f]", r_human, rh_lo, rh_hi)) |>
  knitr::kable(caption = sprintf(paste0(
    "Agreement of the sampled expected value with the logprob expected value and ",
    "with human norms, by number of draws k (full holdout, 18 dimensions; %d random ",
    "k-of-20 resamples, seed 20260604). Reference ceiling: logprob EV vs human ",
    "norms r = %.3f."), n_resamp, ceiling_h))
```

```{r sampling-ksweep-figure, fig.width=7, fig.height=3.6}
library(ggplot2)
plot_df <- bind_rows(
  transmute(ksweep, k, panel = "vs logprob EV",
            r = r_logprob, lo = rlp_lo, hi = rlp_hi),
  transmute(ksweep, k, panel = "vs human norms",
            r = r_human, lo = rh_lo, hi = rh_hi))

p_ksweep <- ggplot(plot_df, aes(k, r)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = .2) +
  geom_line() + geom_point() +
  geom_hline(data = tibble(panel = "vs human norms", y = ceiling_h),
             aes(yintercept = y), linetype = "dashed") +
  facet_wrap(~panel, scales = "free_y") +
  labs(x = "Number of draws (k)", y = "Pearson r",
       title = "Sampled expected value converges to the logprob expected value") +
  theme_minimal(base_size = 11)

ggsave("fig_k_sweep.png", p_ksweep, width = 7, height = 3.6, dpi = 150)
p_ksweep
```

---

## Part B: Logprob SD vs Human Inter-rater SD (logprob models)

**What this shows:** Whether the LLM's distributional uncertainty encodes information beyond its mean rating. Logprob SD is the standard deviation of the probability-weighted rating distribution — large when probability mass is spread across scale values, small when the model is confident. Human inter-rater SD measures participant disagreement. If logprob SD tracks human SD, the model's uncertainty reflects genuine semantic ambiguity rather than noise. Tested per model, then pooled.

### Data

```{r sd-data}
read_dist <- function(rd) {
  p <- file.path(rd, "distributions.csv")
  if (!file.exists(p)) return(NULL)
  read_csv(p, show_col_types = FALSE)
}

dist_all <- bind_rows(lapply(
  runs_tbl$run_dir[runs_tbl$extraction == "logprob"], read_dist
))

# expected (post-transform) keyed by model x word x dimension, holdout only.
exp_map <- dat_lp |> distinct(model_slug, word, dimension, expected)

lp_sd <- dist_all |>
  mutate(rating = llm_prevalence_to_score(rating, dimension)) |>
  inner_join(exp_map, by = c("model_slug", "word", "dimension")) |>
  group_by(model_slug, word, dimension) |>
  summarise(logprob_sd = sqrt(sum(prob * (rating - first(expected))^2)),
            .groups = "drop")

# lexis_long carries a few hundred word x dimension pairs sourced from more than
# one dataset (e.g. "aids"/arousal), each with its own sd. Collapse to one human
# SD per word x dimension so those words are not double-counted.
human_sd <- lexis_long |>
  filter(dimension %in% unique(dat_lp$dimension)) |>
  group_by(word, dimension) |>
  summarise(human_sd = mean(sd, na.rm = TRUE), .groups = "drop") |>
  filter(!is.na(human_sd), is.finite(human_sd))

sd_dat <- lp_sd |>
  inner_join(human_sd, by = c("word", "dimension")) |>
  left_join(dim_groups, by = "dimension") |>
  filter(!is.na(human_sd), !is.na(logprob_sd)) |>
  mutate(group = factor(group,
                        levels = c("Sensory", "Effector", "Affective", "Lexical")))
```

### Per-model per-dimension correlations

Pearson r between logprob SD and human inter-rater SD. Positive = the model is more uncertain on words humans also disagreed about.

```{r sd-cors}
sd_dat |>
  group_by(Model = model_slug, dimension, dim_label, group) |>
  summarise(r = round(cor(human_sd, logprob_sd), 3), n = n(), .groups = "drop") |>
  select(Model, Dimension = dim_label, Group = group,
         `r (logprob SD ~ human SD)` = r, N = n) |>
  arrange(Model, desc(`r (logprob SD ~ human SD)`)) |>
  knitr::kable()
```

### Pooled mixed model: human SD ~ logprob SD

**What this shows:** Human inter-rater SD as the outcome and logprob SD as the predictor, with random intercept and slope by dimension plus a model random intercept. A significant positive `logprob_sd_z` coefficient indicates model uncertainty tracks human disagreement above and beyond dimension- and model-level differences.

```{r sd-lmer}
sd_dat_sc <- sd_dat |>
  group_by(model_slug, dimension) |>
  mutate(
    logprob_sd_z = as.numeric(scale(logprob_sd)),
    human_sd_z   = as.numeric(scale(human_sd))
  ) |>
  ungroup()

m_sd <- lmer(
  human_sd_z ~ logprob_sd_z + (1 + logprob_sd_z | dimension) + (1 | model_slug),
  data = sd_dat_sc, control = lmer_ctrl
)

print(round(summary(m_sd)$coefficients, 4))
print(VarCorr(m_sd))
```

### Sampling recovers the spread, not just the mean

**What this shows:** The sampling section showed that averaging
temperature-1.0 draws reproduces the logprob *mean* (the expected value, *r* = .997).
Here we check the second moment: does the SD of the same 20 draws reproduce the logprob
*SD*? If so, the full distributional reading — both moments — is available without
logprobs, which is the capstone of the distribution story (Part B in the manuscript
reorg). gpt-4o-mini only (the model with a sampled run); `lp_sd` is the per-item logprob
SD built above, the sampled SD is the per-item SD of the 20 draws.

```{r sampling-sd-recovery}
samp_sd <- read_csv(file.path(sampled_dir, "responses_sampled.csv"),
                    show_col_types = FALSE) |>
  filter(split == "holdout", dimension %in% dim_groups$dimension) |>
  transmute(word, dimension, sampled_sd)

sd_recovery <- lp_sd |>
  filter(model_slug == mini_slug) |>
  inner_join(samp_sd, by = c("word", "dimension")) |>
  filter(!is.na(logprob_sd), !is.na(sampled_sd))

cat(sprintf("n cells = %d\n", nrow(sd_recovery)))
cat(sprintf("OVERALL r(sampled SD, logprob SD) = %.3f\n",
            cor(sd_recovery$sampled_sd, sd_recovery$logprob_sd)))

sd_recovery |>
  group_by(dimension) |>
  summarise(r = round(cor(sampled_sd, logprob_sd), 3), n = n(), .groups = "drop") |>
  left_join(dim_groups, by = "dimension") |>
  transmute(Group = group, Dimension = dim_label,
            `r (sampled SD ~ logprob SD)` = r, N = n) |>
  arrange(`r (sampled SD ~ logprob SD)`) |>
  knitr::kable(caption = paste0(
    "Sampling recovers the distributional spread: per-dimension r between the SD of ",
    "20 temperature-1.0 draws and the logprob SD (gpt-4o-mini, full holdout)."))
```

---

## Part A: Residualized Efficacy (horse-race methods)

**What this shows:** Whether method differences survive after controlling for the four lexical covariates (concreteness, frequency, log synset count, n_pos). Human scores and each method's predictions are residualized on the covariates within dimension; residual correlations reflect model-specific signal not explained by where in lexical space the rated words fall.

```{r resid-efficacy}
resid_cor_vec <- function(human, pred, conc, freq, log_syn, npos) {
  df <- data.frame(human = human, pred = pred, conc = conc,
                   freq = freq, log_syn = log_syn, npos = npos)
  df <- df[complete.cases(df), ]
  if (nrow(df) < 10) return(NA_real_)
  cor(residuals(lm(human ~ conc + freq + log_syn + npos, data = df)),
      residuals(lm(pred  ~ conc + freq + log_syn + npos, data = df)))
}

resid_tbl <- horse_long |>
  group_by(method, dimension) |>
  group_modify(~ tibble(
    r_raw = cor(.x$human_score, .x$predicted, use = "complete.obs"),
    r_res = resid_cor_vec(.x$human_score, .x$predicted,
                          .x$concreteness_z, .x$freq_z,
                          .x$log_n_synsets_z, .x$n_pos_z)
  )) |>
  ungroup()

resid_tbl |>
  group_by(Method = method) |>
  summarise(
    Raw    = round(mean(r_raw, na.rm = TRUE), 3),
    Resid  = round(mean(r_res, na.rm = TRUE), 3),
    Shrink = round(mean(r_raw, na.rm = TRUE) - mean(r_res, na.rm = TRUE), 3),
    .groups = "drop"
  ) |>
  arrange(desc(Resid)) |>
  knitr::kable(caption = "Raw vs residualized mean r by method")
```

---

## Summary: Confidence Intervals on Key Fixed Effects

**What this shows:** 95% Wald confidence intervals on the focal fixed effects of the three
main mixed models, complementing the *t*/*p* values reported above. Wald intervals are used
(rather than profiled or parametric-bootstrap) because the focal effects are far from any
boundary and the random-effects structure is simple; they are symmetric by construction.
Terms are matched by pattern so the table is robust to lme4's verbatim factor-level labels.

```{r fixed-effect-cis}
pick_ci <- function(m, pat, label) {
  ci   <- confint(m, method = "Wald")
  fe   <- fixef(m)
  term <- grep(pat, names(fe), value = TRUE)[1]
  tibble(Effect = label,
         Coefficient = term,
         Estimate = round(fe[[term]], 4),
         `2.5%`   = round(ci[term, 1], 4),
         `97.5%`  = round(ci[term, 2], 4))
}

bind_rows(
  pick_ci(m_ei,         "^extraction",            "Pooled expected vs integer (abs error)"),
  pick_ci(m_joint,      "BGE-large:concreteness", "BGE-large x concreteness (range-norm error)"),
  pick_ci(m_ev_uniform, "expected.*concreteness", "EV x concreteness (range-norm error)"),
  pick_ci(m_ev_uniform, "expected.*freq",         "EV x frequency (range-norm error)"),
  pick_ci(m_sd,         "logprob_sd",             "Pooled logprob SD -> human inter-rater SD")
) |>
  knitr::kable(caption = "95% Wald CIs on focal mixed-model fixed effects")
```
