---
title: "Failure-Mode Axis Probe"
subtitle: "Exploratory supervision-confound check"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

## Purpose

This is a companion analysis for the failure-mode frame. It tests whether the
embedding advantage over elicitation can be explained entirely by supervision.
The probe uses embedding geometry with no fitted regression weights: for each
dimension, training words define a low-to-high axis, and holdout words are
projected onto that axis.

The analysis is exploratory and should not be treated as part of the paper until
the results are inspected.

```{r libraries}
suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(tidyr)
  library(purrr)
  library(jsonlite)
  library(ggplot2)
})
```

```{r paths}
find_pkg_root <- function(start = getwd()) {
  cur <- normalizePath(start, mustWork = TRUE)
  repeat {
    if (file.exists(file.path(cur, "DESCRIPTION"))) return(cur)
    parent <- dirname(cur)
    if (identical(parent, cur)) break
    cur <- parent
  }
  stop("Could not find package root.", call. = FALSE)
}

base_dir <- find_pkg_root()
out_dir  <- file.path(base_dir, "analysis", "output")
runs_dir <- file.path(out_dir, "runs")

core_long_path <- file.path(out_dir, "core_vocab_long.csv")
ridge_path     <- file.path(out_dir, "embedding_model_predictions.csv")
bge_path       <- file.path(out_dir, "bge_large_embeddings.csv")

axis_probe_path         <- file.path(out_dir, "failure_mode_axis_probe.csv")
axis_probe_summary_path <- file.path(out_dir, "failure_mode_axis_probe_summary.csv")
axis_probe_focus_path   <- file.path(out_dir, "failure_mode_axis_probe_focus.csv")

stopifnot(
  file.exists(core_long_path),
  file.exists(ridge_path),
  file.exists(bge_path)
)
```

```{r helpers}
`%||%` <- function(x, y) if (is.null(x)) y else x

safe_cor <- function(x, y) {
  ok <- complete.cases(x, y)
  if (sum(ok) < 10) return(NA_real_)
  suppressWarnings(cor(x[ok], y[ok]))
}

l2_normalize <- function(mat) {
  mat <- as.matrix(mat)
  denom <- sqrt(rowSums(mat * mat))
  denom[denom == 0 | is.na(denom)] <- NA_real_
  mat / denom
}

load_rda_object <- function(path, object_name = NULL) {
  env <- new.env(parent = emptyenv())
  nm <- load(path, envir = env)
  if (!is.null(object_name)) return(get(object_name, envir = env))
  get(nm[[1]], envir = env)
}

read_bge <- function(path) {
  raw <- read_csv(path, show_col_types = FALSE)
  mat <- as.matrix(raw[, -1])
  rownames(mat) <- raw$word
  mat
}

probe_dimension <- function(emb, dim_df, pole_n) {
  dim_df <- dim_df |>
    filter(word %in% rownames(emb)) |>
    arrange(score, word)

  train_df <- dim_df |> filter(split == "train")
  hold_df  <- dim_df |> filter(split == "holdout")

  if (nrow(train_df) < 2 * pole_n || nrow(hold_df) < 10) {
    return(tibble(
      word = hold_df$word,
      observed = hold_df$score,
      axis_score = NA_real_
    ))
  }

  low_words  <- head(train_df$word, pole_n)
  high_words <- tail(train_df$word, pole_n)

  low_vec  <- colMeans(emb[low_words, , drop = FALSE], na.rm = TRUE)
  high_vec <- colMeans(emb[high_words, , drop = FALSE], na.rm = TRUE)
  axis     <- high_vec - low_vec

  tibble(
    word = hold_df$word,
    observed = hold_df$score,
    axis_score = as.numeric(emb[hold_df$word, , drop = FALSE] %*% axis)
  )
}

llm_prevalence_to_score <- function(x, dimension) {
  pknown <- pmin(pmax((x - 1) / 9, 0), 1)
  score  <- qnorm(0.005 + 0.99 * pknown)
  ifelse(dimension == "prevalence_score", score, x)
}
```

## Data

```{r load-core}
core_long <- read_csv(core_long_path, show_col_types = FALSE) |>
  filter(dimension != "prevalence_score")

dimensions <- sort(unique(core_long$dimension))

dim_labels <- tibble::tribble(
  ~dimension,                  ~group,       ~dim_label,
  "lancaster_auditory",        "Sensory",    "auditory",
  "lancaster_visual",          "Sensory",    "visual",
  "lancaster_haptic",          "Sensory",    "haptic",
  "lancaster_interoceptive",   "Sensory",    "interoceptive",
  "lancaster_gustatory",       "Sensory",    "gustatory",
  "lancaster_olfactory",       "Sensory",    "olfactory",
  "lancaster_foot_leg",        "Effector",   "foot/leg",
  "lancaster_hand_arm",        "Effector",   "hand/arm",
  "lancaster_head",            "Effector",   "head",
  "lancaster_mouth",           "Effector",   "mouth",
  "lancaster_torso",           "Effector",   "torso",
  "valence",                   "Affective",  "valence",
  "arousal",                   "Affective",  "arousal",
  "dominance",                 "Affective",  "dominance",
  "socialness",                "Lexical",    "socialness",
  "aoa",                       "Lexical",    "AoA",
  "concreteness",              "Lexical",    "concreteness",
  "iconicity",                 "Lexical",    "iconicity"
)
```

```{r load-embeddings}
embeddings <- list(
  bge_large = read_bge(bge_path)
)


embeddings <- lapply(embeddings, l2_normalize)

tibble(
  embedding = names(embeddings),
  n_words = map_int(embeddings, nrow),
  n_dims = map_int(embeddings, ncol)
) |>
  knitr::kable(caption = "Embeddings available for axis probes")
```

```{r load-llm}
discover_run <- function(rd) {
  meta_path <- file.path(rd, "run_meta.json")
  if (!file.exists(meta_path)) return(NULL)
  m <- jsonlite::fromJSON(meta_path)
  responses_path <- file.path(rd, m$responses_file)
  if (!file.exists(responses_path)) return(NULL)
  tibble(
    run_dir = rd,
    run_id = m$run_id,
    model_slug = m$model_slug,
    provider = m$provider %||% NA_character_,
    extraction = m$extraction,
    responses_file = m$responses_file
  )
}

runs_tbl <- list.dirs(runs_dir, recursive = FALSE) |>
  lapply(discover_run) |>
  bind_rows() |>
  filter(extraction != "sampled") |>
  arrange(extraction, model_slug)

read_resp <- function(rd, rf) {
  read_csv(file.path(rd, rf), show_col_types = FALSE)
}

llm_resp <- bind_rows(Map(read_resp, runs_tbl$run_dir, runs_tbl$responses_file)) |>
  filter(split == "holdout", dimension %in% dimensions) |>
  mutate(
    response_integer = llm_prevalence_to_score(integer_rating, dimension)
  )

llm_cors <- llm_resp |>
  filter(!is.na(response_integer)) |>
  group_by(dimension, method = paste0(model_slug, " top-token")) |>
  summarise(
    llm_r = safe_cor(human_score, response_integer),
    .groups = "drop"
  )

featured_llm_method <- if ("gpt-4.1 top-token" %in% llm_cors$method) {
  "gpt-4.1 top-token"
} else {
  llm_cors |>
    group_by(method) |>
    summarise(mean_r = mean(llm_r, na.rm = TRUE), .groups = "drop") |>
    slice_max(mean_r, n = 1, with_ties = FALSE) |>
    pull(method)
}

cat("Featured elicitation comparator:", featured_llm_method, "\n")
```

```{r load-ridge}
ridge_cors <- read_csv(ridge_path, show_col_types = FALSE) |>
  filter(split == "holdout", dimension %in% dimensions) |>
  group_by(dimension, embedding) |>
  summarise(
    ridge_r = safe_cor(observed, predicted),
    .groups = "drop"
  )

featured_ridge <- if ("bge_large" %in% ridge_cors$embedding) {
  "bge_large"
} else {
  ridge_cors |>
    group_by(embedding) |>
    summarise(mean_r = mean(ridge_r, na.rm = TRUE), .groups = "drop") |>
    slice_max(mean_r, n = 1, with_ties = FALSE) |>
    pull(embedding)
}

cat("Featured supervised embedding comparator:", featured_ridge, "\n")
```

## Axis Probe

```{r run-axis-probes}
pole_ns <- c(20L, 30L, 50L, 100L)
primary_pole_n <- 50L

axis_scores <- bind_rows(lapply(names(embeddings), function(embedding_name) {
  emb <- embeddings[[embedding_name]]

  bind_rows(lapply(pole_ns, function(pole_n) {
    bind_rows(lapply(dimensions, function(dimension_name) {
      dim_df <- core_long |> filter(dimension == dimension_name)

      probe_dimension(emb, dim_df, pole_n) |>
        mutate(
          dimension = dimension_name,
          embedding = embedding_name,
          pole_n = pole_n,
          .before = 1
        )
    }))
  }))
}))

axis_summary <- axis_scores |>
  group_by(dimension, embedding, pole_n) |>
  summarise(
    axis_r = safe_cor(observed, axis_score),
    n_holdout = sum(complete.cases(observed, axis_score)),
    .groups = "drop"
  ) |>
  left_join(dim_labels, by = "dimension")

readr::write_csv(axis_scores, axis_probe_path)
readr::write_csv(axis_summary, axis_probe_summary_path)

axis_summary |>
  filter(pole_n == primary_pole_n) |>
  arrange(desc(axis_r)) |>
  mutate(axis_r = round(axis_r, 3)) |>
  select(group, dim_label, embedding, pole_n, axis_r, n_holdout) |>
  knitr::kable(caption = "Axis-probe holdout correlations, primary pole size n = `r primary_pole_n`")
```

## Comparison To Elicitation And Ridge

```{r comparison}
comparison <- axis_summary |>
  filter(pole_n == primary_pole_n) |>
  left_join(
    llm_cors |> filter(method == featured_llm_method),
    by = "dimension"
  ) |>
  left_join(
    ridge_cors |> filter(embedding == featured_ridge) |>
      select(dimension, ridge_embedding = embedding, ridge_r),
    by = "dimension"
  ) |>
  mutate(
    axis_minus_llm = axis_r - llm_r,
    ridge_minus_llm = ridge_r - llm_r,
    probe_beats_llm = axis_minus_llm > 0
  )

readr::write_csv(comparison, axis_probe_focus_path)

comparison |>
  arrange(desc(axis_minus_llm)) |>
  mutate(across(c(axis_r, llm_r, ridge_r, axis_minus_llm, ridge_minus_llm), ~ round(.x, 3))) |>
  select(group, dim_label, embedding, axis_r, method, llm_r,
         ridge_embedding, ridge_r, axis_minus_llm, ridge_minus_llm,
         probe_beats_llm) |>
  knitr::kable(caption = "Primary comparison: axis probe vs elicitation and supervised ridge")
```

```{r headline-focus}
headline_dimensions <- c(
  "lancaster_auditory",
  "lancaster_interoceptive",
  "lancaster_mouth",
  "lancaster_haptic",
  "lancaster_visual",
  "dominance"
)

comparison |>
  filter(dimension %in% headline_dimensions) |>
  arrange(match(dimension, headline_dimensions), embedding) |>
  mutate(across(c(axis_r, llm_r, ridge_r, axis_minus_llm, ridge_minus_llm), ~ round(.x, 3))) |>
  select(group, dim_label, embedding, axis_r, method, llm_r,
         ridge_embedding, ridge_r, axis_minus_llm, ridge_minus_llm,
         probe_beats_llm) |>
  knitr::kable(caption = "Headline failure-mode dimensions")
```

## Sensitivity To Pole Size

```{r pole-sensitivity}
axis_summary |>
  filter(embedding == "bge_large") |>
  ggplot(aes(x = factor(pole_n), y = axis_r, group = dimension)) +
  geom_line(alpha = 0.35) +
  geom_point(size = 1.8) +
  facet_wrap(~ group, scales = "free_y") +
  labs(
    x = "Pole words per side",
    y = "Holdout r",
    title = "BGE axis-probe sensitivity to pole size"
  ) +
  theme_minimal(base_size = 12)
```

```{r comparison-plot}
plot_dat <- comparison |>
  filter(embedding == "bge_large") |>
  transmute(
    dimension, group, dim_label,
    `Axis probe` = axis_r,
    `LLM elicitation` = llm_r,
    `BGE ridge` = ridge_r
  ) |>
  pivot_longer(
    cols = c(`Axis probe`, `LLM elicitation`, `BGE ridge`),
    names_to = "method_family",
    values_to = "r"
  ) |>
  mutate(dim_label = reorder(dim_label, r, FUN = max, na.rm = TRUE))

ggplot(plot_dat, aes(x = dim_label, y = r, color = method_family)) +
  geom_point(position = position_dodge(width = 0.55), size = 2.2) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Holdout r",
    color = NULL,
    title = "Unsupervised axis probe compared with elicitation and ridge"
  ) +
  theme_minimal(base_size = 12)
```

## Readout

The comparison that matters is the axis probe versus elicitation on dimensions
where supervised embeddings beat elicitation. The ridge model is expected to be
stronger than the axis probe because ridge fits a supervised mapping. A weaker
axis probe can still answer the supervision-confound question if it recovers a
dimension that elicitation misses.

## Paired Bootstrap: Probe vs. Elicitation

```{r bootstrap-probe-vs-llm}
set.seed(20260604)
n_boot <- 2000

probe_beat_dims <- comparison |>
  filter(probe_beats_llm, embedding == "bge_large", pole_n == primary_pole_n) |>
  pull(dimension)

featured_model_slug <- sub(" top-token$", "", featured_llm_method)

boot_data <- axis_scores |>
  filter(
    embedding == "bge_large",
    pole_n == primary_pole_n,
    dimension %in% probe_beat_dims
  ) |>
  inner_join(
    llm_resp |>
      filter(model_slug == featured_model_slug) |>
      select(word, dimension, llm_score = response_integer),
    by = c("word", "dimension")
  ) |>
  filter(complete.cases(observed, axis_score, llm_score))

bootstrap_dim <- function(df, n_boot) {
  n <- nrow(df)
  vapply(seq_len(n_boot), function(i) {
    idx <- sample(n, n, replace = TRUE)
    safe_cor(df$observed[idx], df$axis_score[idx]) -
      safe_cor(df$observed[idx], df$llm_score[idx])
  }, numeric(1))
}

boot_results <- boot_data |>
  group_by(dimension) |>
  group_modify(~ {
    diffs <- bootstrap_dim(.x, n_boot)
    tibble(
      observed_diff = safe_cor(.x$observed, .x$axis_score) -
                      safe_cor(.x$observed, .x$llm_score),
      ci_lo    = quantile(diffs, 0.025),
      ci_hi    = quantile(diffs, 0.975),
      p_gt_zero = mean(diffs > 0)
    )
  }) |>
  ungroup() |>
  left_join(dim_labels, by = "dimension") |>
  arrange(desc(observed_diff))

boot_results |>
  mutate(
    across(c(observed_diff, ci_lo, ci_hi), ~ round(.x, 3)),
    p_gt_zero = round(p_gt_zero, 3),
    clean_win = ci_lo > 0
  ) |>
  select(group, dim_label, observed_diff, ci_lo, ci_hi, p_gt_zero, clean_win) |>
  knitr::kable(
    caption = paste0(
      "Paired bootstrap: axis probe minus ", featured_llm_method,
      " (", n_boot, " resamples, seed 20260604)"
    )
  )
```

```{r save-bootstrap}
readr::write_csv(
  boot_results |>
    left_join(dim_labels, by = "dimension") |>
    mutate(clean_win = ci_lo > 0),
  file.path(out_dir, "failure_mode_probe_bootstrap.csv")
)
```

